Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets

### 2021-09-19 publications/iaai21/dora_outlier_detection_iaai.tex
Trim text to 6 pages. #42

In astrophysics, 
outlier detection methods have been used to identify astrophysical
objects with unique characteristics~\cite{xiong2010anomaly,
storey2020anomaly,hayat2021self} as well as data or modeling 
artifacts in astronomical surveys~\citep{wagstaff:des-anom20,
lochner2021astronomaly}. 
Example outlier detection applications in 
Earth science include detecting anomalous objects or materials
~\cite{chang2002anomaly,zhou2016novel}, data artifacts or noise 
~\cite{liu2017unsupervised,alvera2012outlier}, change~\cite{zhou2016novel,
touati2020anomaly}, and ocean extremes~\cite{prochaska2021deep}. 
Planetary science applications have mostly focused 
on prioritizing samples with novel geologic or geochemical features for
follow-up targeting or analysis~\cite{kerner2020analysis,kerner2020comparison,
wagstaff:demud13}.
%
These examples show the benefit of applying outlier detection methods in
a variety of real-world science use cases. However, the effort required to 
apply and evaluate the many available algorithms is non-trivial and can 
be daunting for non-ML experts, thus impeding the uptake of outlier
detection methods in science applications.  
There is a need for tools that make it easier for domain scientists to apply
outlier detection methods as well as compare results across datasets.
While there have been some efforts to develop tools for facilitating the
application of outlier detection methods~\cite{zhao2019pyod}, they cover
limited data formats and algorithms. The aim of DORA is to fill the need 
for tools that faciliate application, evaluation, and interpretation of outlier
detection methods.

\section{Methods}

%\subsection{DORA configurable pipeline} 
Figure \ref{fig:dora} illustrates the architecture of DORA. It includes 
data loading, feature extraction, outlier ranking, results organization, and 
results visualizaiton modules. The DORA pipeline is implemented in Python 3.
In order to improve the readability and execution speed of the code, we adopted
object-oriented and functional programming practices.  We designed DORA to be 
readily extensible to support additional data types or formats, outlier detection 
algorithms, and results organization or visualization methods by writing new 
modules that follow the DORA API. Experimental settings are controlled by 
configuration files. In a configuration file, users can specify the input data, 
feature extraction methods, normalization method, outlier ranking methods, and 
results organization methods.

@@

\subparagraph{Reconstruction error.}
Principal component analysis (PCA) has been used for outlier detection by
scoring samples using the reconstruction error (here, the L2 norm)
between inputs and their inverse
 transformation from the principal subspace~\citep{kerner2020comparison}.
The DEMUD algorithm~\citep{wagstaff:demud13} differs from other
outlier ranking methods: instead of independently scoring all
observations, DEMUD incrementally identifies the most unusual
remaining item, then incorporates it into the model of ``known''
(non-outlier) observations before selecting the next most unusual
item.  DEMUD's goal is to identify diverse outliers and avoid
redundant selections.  Once an outlier is encountered, repeated
occurrences of that outlier are deprioritized.  Methods that score
outliers independently maximize coverage of outliers, while DEMUD
maximizes the fast discovery of distinct outlier types.

\subparagraph{Distance.}
 The 
Reed-Xiaoli (RX) detector is commonly used for anomaly detection in
multispectral and hyperspectral remote sensing; RX uses the Mahalanobis
 distance between a sample and a background dataset mean and covariance
 to score samples~\citep{reed1990adaptive}. The local variant of RX (Local
 RX or LRX) can be used for image or raster data types and scores each pixel in
  an image with respect to a window ``ring'' of pixels surrounding 
  it~\citep{molero2013analysis}. 
  LRX requires two parameters that define the ring: an outer window which 
  defines the size of the window surrounding the pixel and an inner window
   which defines the size of the window around the target pixel to exclude from
   the background distribution.

@@

\paragraph{Results interpretation.}
Each of the outlier ranking algorithms returns an array containing the sample
index, outlier score, and selection index (index of each sample after sorting
by outlier score). DORA provides organization and visualization modules
intended to help users interpret and make decisions based on these outputs.
The simplest
 module saves a CSV of the samples sorted by their outlier score 
 (i.e., selection order). Clustering the top
  $N$ outlier selections can enable users to investigate the different types of 
 outliers that might be present in the dataset; this could be especially useful
 for separating outliers caused by noise or data artifacts vs. scientifically 
 interesting samples. We implemented the K-means and self-organizing maps 
 (SOMs) algorithms for clustering the top-N outliers. For use cases in which an
 evaluation dataset containing known outliers is available, we provide a module
 to assess how well algorithm selections correlate with known outliers. This is
 done by plotting the number of true outliers vs. number of selections made. 
We provide a module for plotting histograms of outlier scores to visualize the
 distribution of scores in the dataset (which may be, e.g., multimodal or 
 long-tailed).

@@

We developed a desktop application to easily visualize DORA results with the
Electron application framework and React frontend library. The application loads
the DORA configuration file to locate the dataset and result CSVs. Then,
it displays the ranked samples and their scores in a table sorted by their
selection order. This allows for fast and easy comparison of the results of
different methods. Figure \ref{fig:doravis} shows a screenshot of the
``Aggregate Table'' view, which displays all results from different algorithms
side-by-side.

\section{Datasets}
\begin{table}
  \caption{Number of samples in the training and test sets for each dataset,
  ordered by dataset size.}
  \label{tab:datasets}
  \centering
  \begin{tabular}{l|c|cc}
    \hline
    Dataset & Training & \multicolumn{2}{c}{Test}\\
     & Unlabeled &  Outliers &  Inliers \\

@@

We constructed three datasets to evaluate the utility of DORA and 
algorithm performance for a variety of scientific domains
 (astrophysics, planetary science, and Earth science). We also included a 
 benchmark dataset that uses MNIST and Fashion-MNIST. Table \ref{tab:datasets}
 summarizes the number of unlabeled samples used for training evaluation for
 each dataset. We describe each dataset in detail below.

@@

\subsubsection{Planetary: targets in Mars rover images.}
Mars exploration is fundamentally an exercise in discovery with the
goal of increasing our understanding of the history, evolution,
composition, and currently active processes on Mars.  Outliers
identified in Mars observations can inspire new discoveries and inform
the choice of which areas merit follow-up or deeper
investigation~\cite{kerner2020comparison,wagstaff:rover-novelty20}.
We collected \num{72} images from the Navigation camera (Navcam) 
on the Mars Science Laboratory (MSL) rover
and employed the Rockster~\cite{burl:rockster16}
method currently used by onboard rover software to identify candidate rock
targets with an area of at least 100 pixels, yielding \num{1050}
targets.  We cropped out a \num{64}$\times$\num{64} pixel image
centered on each target.

@@

\begin{table}
  \caption{Precision at $N=n_{outlier}$ for three applications and F-MNIST
   benchmark; the best result for each data set is in bold.}
  \label{tab:faw_results}
  \centering
  \begin{tabular}{l|cccc}
    \hline
    Algorithm & Astro. & Earth & Planetary & F-MNIST \\
    \hline
    PCA               & \textbf{0.42} & 0.41 & 0.44 & \textbf{0.84} \\
    DEMUD             & ---- & 0.41 & 0.44 & \textbf{0.84} \\
    \hline
    RX                & 0.40 & 0.43 & ---- & 0.82 \\
    LRX               & ----& ---- & 0.33 & 0.56 \\
    \hline
    Iso. Forest  & 0.34 & 0.46 & \textbf{0.56} & 0.74 \\
    \hline
    PAE               & 0.35 & 0.41 & 0.44 & 0.83 \\
    Neg. Sampling & 0.32 & \textbf{0.49} & 0.33 & 0.43 \\
    \hline
    Random            & 0.25 & 0.32 & 0.14 & 0.50 \\
    \hline
  \end{tabular}
\end{table}

\section{Results}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/planetary_combined_plot.pdf}
    \caption{Number of true outliers ranked in top $N$ selections for 
    planetary rover dataset.}
    \label{fig:planetary_results}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/faw_combined_plot.pdf}
    \caption{Number of true outliers ranked in top $N$ selections for 
    Earth science dataset.}
    \label{fig:faw_results}
\end{figure}
The experimental setup for each dataset was to fit or train a model for
each ranking algorithm using a larger unlabeled dataset and then apply 
the models to compute the outlier scores for a smaller test dataset for which
labels of known outliers were available (Table \ref{tab:datasets}). 
For each test set, we created a plot of the number of known/true outliers 
detected out of the top $N$ selections. We also 
reported the Mean Discovery Rate (MDR) in the legend
for each algorithm to give a quantitative comparison of the curves that could
be comparable across the datasets. We defined MDR as:
\begin{equation}
MDR = \frac{\sum^{N_s}_{i=1} n_i}{\sum^{N_s}_{i=1} s_i}
\end{equation}
where $i \in [1, N_s]$ is the selection index,
 $N_s$ is the total number of selections,
 $s_i$ is the number of selections made up to index $i$,
 and $n_i$ is the number of known outliers (true positives)
among $s_i$ selections.
We also reported the precision at $N=n_{outlier}$ for each test set where
$n_{outlier}$ is the number of known outliers, i.e.,  the precision obtained 
when the number of selections is the same as the total number of outliers. 
Precision at $N$ is the number of true outliers divided by the number of 
selections $N$~\citep{campos2016evaluation}.
Table~\ref{tab:faw_results} compares the precision at $N=n_{outlier}$ for
each dataset and ranking algorithm. 
We calculated a random selection baseline which we refer to as ``Theoretical
Random'' using the expected value of $n_i$ for $i$ random selections:
\begin{align}
\mathbf{E}[n_i, i \in[1, N_s]] &= \frac{\sum^i_{j=0} {n_{outlier}\choose j} {D-n_{outlier}\choose i-j} j}{{D\choose i}} \\
&= \frac{n_{outlier}i}{D}
\end{align}

@@

For the Earth dataset, the negative sampling algorithm had the best performance
in both metrics. DEMUD, PCA, and PAE tied for the lowest precision at 
$N=n_{outlier}$ while DEMUD and PCA tied for the lowest
MDR (Figure \ref{fig:faw_results}). We did not evaluate LRX for this time 
series dataset because LRX can only be applied to gridded image or raster 
data types.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/des_combined_plot.pdf}
    \caption{Number of true outliers ranked in top $N$ selections for 
    the astrophysics (DES) dataset.}
    \label{fig:des_results}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/fmnist_combined_plot.pdf}
    \caption{Number of true outliers ranked in top $N$ selections for 
    F-MNIST dataset.}
    \label{fig:fmnist_results}
\end{figure}

@@

\paragraph{Algorithm performance.} 
There was no one algorithm that 
 had the best performance across all four datasets. PCA had
the best performance for the astrophysics and F-MNIST datasets, 
while negative sampling and isolation forest was best for
the Earth science and planetary datasets respectively. This illustrates
the importance of including a diverse set of algorithms and tools for
easily inter-comparing them in DORA, since the best
algorithm will vary for different datasets. The purpose of this study
was to demonstrate how DORA could be used to facilitate 
outlier detection experiments and compare results across datasets from 
different domains. Thus we chose hyperparameters based on common 
practice and sense and left hyperparameter tuning, which would likely 
improve results for each dataset but would require
creation of additional validation sets, for future work.

@@

\todo{Acknowledge DES?}
We thank the UN Food and Agriculture Organization (FAO) and Dr. Ritvik Sahajpal
(University of Maryland/NASA Harvest) for providing the fall army worm ground
 observation dataset.

>>>>

In astrophysics, 
outlier detection methods have been used to identify astrophysical
objects with unique characteristics~\cite{xiong2010anomaly,
storey2020anomaly,hayat2021self} as well as data or modeling 
artifacts in astronomical surveys~\citep{wagstaff:des-anom20,
lochner2021astronomaly}. 
Example outlier detection applications in 
Earth science include detecting anomalous objects or materials
~\cite{chang2002anomaly,zhou2016novel}, data artifacts or noise 
~\cite{liu2017unsupervised,alvera2012outlier}, change~\cite{zhou2016novel,
touati2020anomaly}, and ocean extremes~\cite{prochaska2021deep}. 
Planetary science applications have mostly focused 
on prioritizing samples with novel geologic or geochemical features for
follow-up targeting or analysis~\cite{kerner2020analysis,kerner2020comparison,
wagstaff:demud13}.
%
These examples show the benefit of applying outlier detection methods in
a variety of real-world science use cases. However, the effort required to 
apply and evaluate the many available algorithms is non-trivial and can 
be daunting for non-ML experts, thus impeding the uptake of outlier
detection methods in science applications.  
There is a need for tools that make it easier for domain scientists to apply
outlier detection methods as well as compare results across datasets.
While there have been some efforts to develop tools for facilitating the
application of outlier detection methods~\cite{zhao2019pyod}, they cover
limited data formats and algorithms. DORA aims to fill the need 
for tools that faciliate application, evaluation, and interpretation of outlier
detection methods.

\section{Methods}
%\subsection{DORA configurable pipeline} 
Figure \ref{fig:dora} illustrates the architecture of DORA including 
data loading, feature extraction, outlier ranking, and results organization
and visualization modules. 
In order to improve the readability and execution speed of the code, we adopted
object-oriented and functional programming practices.  We designed DORA to be 
readily extensible to support additional data types or formats, outlier detection 
algorithms, and results organization or visualization methods by writing new 
modules that follow the DORA API. Experimental settings are controlled by 
configuration files in which users can specify the input data, 
feature extraction methods, normalization method, outlier ranking methods, and 
results organization methods. 
DORA is implemented in Python 3.

@@

\subparagraph{Reconstruction error.}
Principal component analysis (PCA) has been used for outlier detection by
scoring samples using the reconstruction error (here, the L2 norm)
between inputs and their inverse
 transformation from the principal subspace~\citep{kerner2020comparison}.
 DEMUD~\citep{wagstaff:demud13} differs from other
outlier ranking methods: instead of independently scoring all
observations, DEMUD incrementally identifies the most unusual
remaining item, then incorporates it into the model of ``known''
(non-outlier) observations before selecting the next most unusual
item.  DEMUD's goal is to identify diverse outliers and avoid
redundant selections.  Once an outlier is found, repeated
occurrences of that outlier are deprioritized.  Methods that score
samples independently maximize coverage of outliers, while DEMUD
maximizes fast discovery of distinct outlier types.

\subparagraph{Distance.}
 The 
Reed-Xiaoli (RX) detector is commonly used for anomaly detection in
multispectral and hyperspectral remote sensing. RX scores samples using
 the Mahalanobis
 distance between a sample and a background mean and covariance
 ~\citep{reed1990adaptive}. The local variant of RX (Local
 RX or LRX) can be used for image or raster data and scores each pixel in
  an image with respect to a window ``ring'' of pixels surrounding 
  it~\citep{molero2013analysis}. 
  LRX requires two parameters to define the size of the outer window 
  surrounding the pixel and the inner window
   around the target pixel to exclude from the background distribution.

@@

\paragraph{Results interpretation.}
Each of the outlier ranking algorithms returns an array containing the sample
index, outlier score, and selection index (index after sorting
by outlier score). DORA provides organization and visualization modules
intended to help users interpret and make decisions based on these outputs.
The simplest
 module saves a CSV of the samples sorted by their outlier score 
 (i.e., selection order). Clustering the top
  $N$ outlier selections can enable users to investigate the different types of 
 outliers that might be present in the dataset; this could be especially useful
 for separating outliers caused by noise or data artifacts vs. scientifically 
 interesting samples. We implemented the K-means and self-organizing maps 
 (SOMs) algorithms for clustering the top-N outliers. For use cases in which an
 evaluation dataset containing known outliers is available, we provide a module
 to assess how well algorithm selections correlate with known outliers. This is
 done by plotting the number of known outliers vs. number of selections made. 
We provide a module for plotting histograms of outlier scores to visualize the
 distribution of scores in the dataset (which may be, e.g., multimodal or 
 long-tailed). 
We developed a desktop application to easily visualize DORA results with the
Electron application framework and React frontend library. 
This enables fast and easy comparison of the results from
different methods. Figure \ref{fig:doravis} shows a screenshot of the
``Aggregate Table'' view, which displays all algorithm results 
side-by-side.

%We developed a desktop application to easily visualize DORA results with the
%Electron application framework and React frontend library. The application loads
%the DORA configuration file to locate the dataset and result CSVs. Then,
%it displays the ranked samples and their scores in a table sorted by their
%selection order. This allows for fast and easy comparison of the results of
%different methods. Figure \ref{fig:doravis} shows a screenshot of the
%``Aggregate Table'' view, which displays all results from different algorithms
%side-by-side.

@@

\section{Datasets}
We constructed three datasets to evaluate the utility of DORA and 
algorithm performance for a variety of scientific domains
 (astrophysics, planetary science, and Earth science). We also included a 
 benchmark dataset that uses MNIST and Fashion-MNIST. Table \ref{tab:datasets}
 summarizes the number of unlabeled samples used for training evaluation for
 each dataset. We describe each dataset in detail below.

 \begin{table}
  \caption{Number of samples in the training and test sets for each dataset,
  ordered by dataset size.}
  \label{tab:datasets}
  \centering
  \begin{tabular}{l|c|cc}
    \hline
    Dataset & Training & \multicolumn{2}{c}{Test}\\
     & Unlabeled &  Outliers &  Inliers \\

@@

@@

\subsubsection{Planetary: targets in Mars rover images.}
Mars exploration is fundamentally an exercise in discovery with the
goal of increasing our understanding of Mars's history, evolution,
composition, and currently active processes.  Outliers
identified in Mars observations can inspire new discoveries and inform
the choice of which areas merit follow-up or deeper
investigation~\cite{kerner2020comparison,wagstaff:rover-novelty20}.
We collected \num{72} images from the Navigation camera (Navcam) 
on the Mars Science Laboratory (MSL) rover
and employed Rockster~\cite{burl:rockster16}
(currently used by onboard rover software) to identify candidate rock
targets with an area of at least 100 pixels, yielding \num{1050}
targets.  We cropped out a \num{64}$\times$\num{64} pixel image
centered on each target.

@@

\begin{table}
  \caption{Precision at $N=n_{outlier}$ for four datasets; the best result for each data set is in bold.}
  \label{tab:faw_results}
  \centering
  \begin{tabular}{l|cccc}
    \hline
    Algorithm & Astro. & Earth & Planetary & F-MNIST \\
    \hline
    PCA               & \textbf{0.42} & 0.41 & 0.44 & \textbf{0.84} \\
    DEMUD             & ---- & 0.41 & 0.44 & \textbf{0.84} \\
    \hline
    RX                & 0.40 & 0.43 & ---- & 0.82 \\
    LRX               & ----& ---- & 0.33 & 0.56 \\
    \hline
    Iso. Forest  & 0.34 & 0.46 & \textbf{0.56} & 0.74 \\
    \hline
    PAE               & 0.35 & 0.41 & 0.44 & 0.83 \\
    Neg. Sampling & 0.32 & \textbf{0.49} & 0.33 & 0.43 \\
    \hline
    Random            & 0.25 & 0.32 & 0.14 & 0.50 \\
    \hline
  \end{tabular}
\end{table}

\section{Results}
\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{figures/planetary_combined_plot.pdf}
    \caption{Number of known outliers ranked in top $N$ selections for 
    planetary rover dataset.}
    \label{fig:planetary_results}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{figures/faw_combined_plot.pdf}
    \caption{Number of known outliers ranked in top $N$ selections for 
    Earth science dataset.}
    \label{fig:faw_results}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{figures/des_combined_plot.pdf}
    \caption{Number of known outliers ranked in top $N$ selections for 
    the astrophysics (DES) dataset.}
    \label{fig:des_results}
\end{figure}

The experimental setup for each dataset was to fit or train a model for
each ranking algorithm using a larger unlabeled dataset and then apply 
the models to compute the outlier scores for a smaller test dataset for which
labels of known outliers were available (Table \ref{tab:datasets}). 
For each test set, we created a plot of the number of known outliers 
detected out of the top $N$ selections. We also 
reported the Mean Discovery Rate (MDR) in the legend
for each algorithm to give a quantitative comparison of the curves that could
be comparable across the datasets. We defined MDR as:
\begin{equation}
MDR = \frac{\sum^{N_s}_{i=1} n_i}{\sum^{N_s}_{i=1} s_i}
\end{equation}
where $i \in [1, N_s]$ is the selection index,
 $N_s$ is the total number of selections,
 $s_i$ is the number of selections made up to index $i$,
 and $n_i$ is the number of known outliers (true positives)
among $s_i$ selections.
We also reported the precision at $N=n_{outlier}$ for each test set where
$n_{outlier}$ is the number of known outliers, i.e.,  the precision obtained 
when the number of selections is the same as the total number of outliers. 
Precision at $N$ is the number of known outliers divided by the number of 
selections $N$~\citep{campos2016evaluation}.
Table~\ref{tab:faw_results} compares the precision at $N=n_{outlier}$ for
each dataset and ranking algorithm. 
We calculated a random selection baseline which we refer to as ``Theoretical
Random'' using the expected value of $n_i$ for $i$ random selections:
\begin{align}
\mathbf{E}[n_i, i \in[1, N_s]] &= \frac{\sum^i_{j=0} {n_{outlier}\choose j} {D-n_{outlier}\choose i-j} j}{{D\choose i}} \\
&= \frac{n_{outlier}i}{D}
\end{align}

@@

For the Earth dataset, negative sampling had the best performance
in both metrics. DEMUD, PCA, and PAE tied for the lowest precision at 
$N=n_{outlier}$ while DEMUD and PCA tied for the lowest
MDR (Figure \ref{fig:faw_results}). We did not evaluate LRX for this time 
series dataset because LRX can only be applied to gridded image or raster 
data types.

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/fmnist_combined_plot.pdf}
    \caption{Number of known outliers ranked in top $N$ selections for 
    F-MNIST dataset.}
    \label{fig:fmnist_results}
\end{figure}

\paragraph{Algorithm performance.} 
No one algorithm  
 had the best performance across all four datasets. PCA had
the best performance for the astrophysics and F-MNIST datasets, 
while negative sampling and isolation forest was best for
the Earth science and planetary datasets respectively. This illustrates
the importance of including a diverse set of algorithms and tools for
easily inter-comparing them in DORA, since the best
algorithm will vary for different datasets. The purpose of this study
was to demonstrate how DORA could be used to facilitate 
outlier detection experiments and compare results across datasets from 
different domains. Thus we did not perform hyperparameter tuning which
could improve results for each dataset; we leave this for future work.

@@

\todo{Acknowledge DES?}
We thank the UN Food and Agriculture Organization (FAO) and Dr. Ritvik Sahajpal
(University of Maryland/NASA Harvest) for providing the maize ground
 observation dataset.
